Demes: a standard format for demographic models

### 2022-08-03 paper.tex
Rejig popgen model
Move content up from appendix, and try to be clearer on the motivation

\subsection*{Population genetics model}

Demographic models consist of one or more populations (or ``demes'') defined
by their size histories and the time intervals of their existence.
Individuals can move between populations based on their ancestor-descendant
relationships  or by continuous or discrete migration events.
The demographic model itself omits information about genome biology, such as
genome size and architecture, ploidy, and mutation and recombination rates.
It also omits any information about selection operating within populations.
Within populations, allele-frequency dynamics are assumed to
follow the Wright-Fisher model, so that generations are discrete and
non-overlapping and population sizes are independent of mean fitness
(i.e., ``soft'' selection~\citep{christiansen1975hard}).
While these restrictions preclude some population-genetic scenarios of
interest, such as continuous spatial structure
\citep{wright1943isolation,barton2002neutral,barton2010new,
ringbauer2017inferring,battey2020space}
or stochastic population size dynamics
\citep{nunney1993assessing,orr2014population},
many current forwards- and backwards-in-time simulators
make these same assumptions.
The basic assumptions of discrete populations connected by instantaneous or continuous
migrations are also shared by many inference methods
\citep[e.g.,][]{gutenkunst2009inferring,li2011inference,gravel2012population,kamm2017efficient,
jouganous2017inferring,ragsdale2019models,excoffier2021fastsimcoal2}.
Demes therefore serves as ``middleware'' between inference methods and simulation
software, capturing these common assumptions.
Appendix~\ref{sec:appendix-pop-gen-model} provides more details on the
population genetics models supported by Demes.

@@

\subsection{Population dynamics}

Within a population, we assume that allele frequency dynamics can
be described by the Wright-Fisher model.
Briefly, generations are non-overlapping (all parents
reproduce and die simultaneously), and for allele $i$ currently at frequency $p_i$,
its frequency in the next generation (at birth) is expected to be $p_iw_i/\bar{w}$,
where $w_i$ and $\bar{w}$ are the marginal and
mean fitnesses, respectively, properly weighted according to ancestry proportions.
In this framework,
a forward-time simulation of finite populations is equivalent
to multinomial sampling of allele frequencies each
generation (\citet[][pp 29-31]{burger2000-ul}, \citet[][pp 179-181]{crowkimura1970}),
and a backwards-time (coalescent) simulation follows the approximations
described in \citet{tajima1983evolutionary,hudson1983testing}
and \citet[][chapter 3]{wakeley2008-hd}.
Further, this model assumes ``soft'' selection \citep{christiansen1975hard},
meaning that the dynamics of population sizes
changes are independent of the details of individual fitnesses.
As such, this model excludes scenarios such as ``hard selection,''
in which population sizes are dependent on a population's mean fitness, or
stochastic fluctuations in population size, such as interpreting
population sizes as carrying capacities.
Many forwards and backwards time simulators currently implement this model
\citep[e.g.,][]{hudson2002generating,gutenkunst2009inferring,excoffier2011fastsimcoal,kelleher2016efficient,jouganous2017inferring,haller2019slim,thornton2019-nu}.
Each has their own unique method of specifying the details of selection and demography.
The goal of Demes is to provide a common method for the latter task.

>>>>

\subsection*{Population genetics model}
For inference and simulation software to meaning fully interoperate there
must be a shared understanding of what a demographic model \emph{is}.
Population genetics is a large field, and rather than attempting to
capture all possible within- and between- population processes, we have
instead adopted a pragmatic approach of identifying a common set
of assumptions shared by many methods. We outline the processes
and assumptions briefly here and in
Appendix~\ref{sec:appendix-pop-gen-model}.

Demographic models consist of one or more populations (or ``demes'') defined
by their size histories and the time intervals of their existence.
Individuals can move between populations based on their ancestor-descendant
relationships  or by continuous or discrete migration events.
As described in the Scope of the Specification section below, the demographic
model does not, as a deliberate simplification and separation of
duties, include any information about genome biology or selection.
Within a population, we assume Wright-Fisher dynamics
(see Appendix~\ref{sec:appendix-population-dynamics} for more
precise details).

These basic assumptions of discrete Wright-Fisher populations connected by
instantaneous or continuous
migrations are shared by many inference methods
\citep[e.g.,][]{gutenkunst2009inferring,li2011inference,
gravel2012population,kamm2017efficient,
jouganous2017inferring,ragsdale2019models,
excoffier2021fastsimcoal2}, and
forwards- and backwards-time simulators~\citep[e.g.,][]{hudson2002generating,
gutenkunst2009inferring,
excoffier2011fastsimcoal,kelleher2016efficient,
jouganous2017inferring,haller2019slim,thornton2019-nu}.
Demes therefore serves as ``middleware'' between inference methods and simulation
software, capturing these common assumptions.

It is important to note that the goal of describing
the basic population processes precisely is not to be
proscriptive about what methods may or may not use the specification,
but so that we can be clear on what situations we can
expect methods to agree exactly. Arbitrary population
processes
---for example, within-deme
continuous spatial structure
\citep{wright1943isolation,barton2002neutral,barton2010new,
ringbauer2017inferring,battey2020space}---may
be layered on top of this basic description,
but as dynamics diverge from the core assumptions,
then of course we can expect results to differ
accordingly.

@@

\subsection{Population dynamics}
\label{sec:appendix-population-dynamics}
Within a population, we assume that allele frequency dynamics can
be described by the Wright-Fisher model.
Briefly, generations are non-overlapping (all parents
reproduce and die simultaneously), and for allele $i$ currently at frequency $p_i$,
its frequency in the next generation (at birth) is expected to be $p_iw_i/\bar{w}$,
where $w_i$ and $\bar{w}$ are the marginal and
mean fitnesses, respectively, properly weighted according to ancestry proportions.
In this framework,
a forward-time simulation of finite populations is equivalent
to multinomial sampling of allele frequencies each
generation (\citet[][pp 29-31]{burger2000-ul}, \citet[][pp 179-181]{crowkimura1970}),
and a backwards-time (coalescent) simulation follows the approximations
described in \citet{tajima1983evolutionary,hudson1983testing}
and \citet[][chapter 3]{wakeley2008-hd}.
Further, this model assumes ``soft'' selection \citep{christiansen1975hard},
meaning that the dynamics of population sizes
changes are independent of the details of individual fitnesses.
As such, this model excludes scenarios such as ``hard selection,''
in which population sizes are dependent on a population's mean fitness, or
stochastic fluctuations in population size, such as interpreting
population sizes as carrying capacities.
Many forwards and backwards time simulators currently implement this model
\citep[e.g.,][]{hudson2002generating,gutenkunst2009inferring,excoffier2011fastsimcoal,kelleher2016efficient,jouganous2017inferring,haller2019slim,thornton2019-nu}.
